---
title: "CyclisticBikeShare"
author: "Tejaswi"
date: "2022-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This research is for Google Data Analytics Certificate Capstone Project.It was inspired by Kevin Hartman's case study "'Sophisticated, Clear, and Polished': Divvy and Data Visualization" (available here:https://artscience.blog/home/divvy-dataviz-case-study).

## Scenario

I am a junior data analyst working in the marketing analyst team at Cyclistic, a bike share company in chicago. The director of marketing believes the company's future success depends on maximizing the number of annual memberships. Therefore my team wants to understand how casual riders and annual members use cyclistic bikes differently and design a new marketing strategy to convert casual riders into annual members

## Key Stakeholders

* Lily Moreno, Director of Marketing
* Marketing Analytics Team
* Cyclistic Executive Team

## Business Task

To understand how casual riders and annual members use cyclistic bikes differently and design marketing strategy to maximize the number of annual Memberships by converting casual riders into annual members

## Preparing Data

The data has been made available by Motivate International Inc. The data is original and [Licensed](https://ride.divvybikes.com/data-license-agreement)

Citation - The data is available on https://www.divvybikes.com/system-data and is owned by the City of Chicago

#### Installing required Packages

Use install.packages("tidyverse") to install tidyverse
Tidyverse - A system of packages in R with a common design philosophy for data manipulation, exploration and visualization

Lubridate - It is an R package that makes it easier to work with dates and times

ggplot2 - Creates a variety of data visualizations by applying different visual properties to the data variables in R.

```{r cars}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

#### Collecting Data

readr - It is used for importing data

read_csv() -> This function is used to import csv files and reads the data as data frames from it. it is available under readr package.
```{r collecting data}
library(readr)
# Uploading divvy datasets here
Q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
Q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
Q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
Q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

#### Wrangle Data and Combine to a Single File

compare column names of each of the files
While the column names need not be in same order, but they do need to perfectly match before we can use a command to join them into one file.

```{r wrangle data}
colnames(Q2_2019)
colnames(Q3_2019)
colnames(Q4_2019)
colnames(Q1_2020)
```

Rename columns to make them consistent with Q1_2020(as this will be the supposed going forward table design for Divvy)

```{r renaming columns}
Q4_2019 <- rename(Q4_2019,
                  ride_id = trip_id,
                  rideable_type = bikeid,
                  started_at = start_time,
                  ended_at = end_time,
                  start_station_name = from_station_name,
                  start_station_id = from_station_id,
                  end_station_name = to_station_name,
                  end_station_id = to_station_id,
                  member_casual = usertype)
(Q3_2019 <- rename(Q3_2019,
                   ride_id = trip_id,
                   rideable_type = bikeid,
                   started_at = start_time,
                   ended_at = end_time,
                   start_station_name = from_station_name,
                   start_station_id = from_station_id,
                   end_station_name = to_station_name,
                   end_station_id = to_station_id,
                   member_casual = usertype))
Q2_2019 <- rename(Q2_2019,
                  ride_id = "01 - Rental Details Rental ID",
                  rideable_type = "01 - Rental Details Bike ID",
                  started_at = "01 - Rental Details Local Start Time",
                  ended_at = "01 - Rental Details Local End Time",
                  start_station_name = "03 - Rental Start Station Name",
                  start_station_id = "03 - Rental Start Station ID",
                  end_station_name = "02 - Rental End Station Name",
                  end_station_id = "02 - Rental End Station ID",
                  member_casual = "User Type")
str(Q2_2019)
str(Q3_2019)
str(Q4_2019)
str(Q1_2020)
```
Convert ride_id and rideable_type to character so that they can stack correctly

```{r converting data type}
Q4_2019 <- mutate(Q4_2019,
                  ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
Q3_2019 <- mutate(Q3_2019,
                  ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
Q2_2019 <- mutate(Q2_2019,
                  ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
```

stack individual quarters data frames to one data frame
```{r stacking to one data frame}
all_trips <- bind_rows(Q2_2019,Q3_2019,Q4_2019,Q1_2020)
all_trips <- all_trips %>% 
  select(-c(start_lat,start_lng,end_lat,end_lng,birthyear,gender,"01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year","Member Gender","tripduration"))
```

## clean up and add data for analysis

inspect the new table that has been created

```{r inspecting data}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)

```
There are few problems that need to be fixed: they are

* In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.

* The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.

* We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.

* There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.


 In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual". Before 2020, Divvy used different labels for these two types of riders ... we will want to make our dataframe consistent with their current nomenclature
 
```{r renaming values of member_casual}
table(all_trips$member_casual)
all_trips <- all_trips %>% 
  mutate(member_casual = recode(member_casual,"Subscriber" = "member", "Customer" = "casual"))
table(all_trips$member_casual)

```

Add columns that list the date, month, day, and year of each ride. This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level

```{r adding columns that list date,month,year and date}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date),"%m")
all_trips$day <- format(as.Date(all_trips$date),"%d")
all_trips$year <- format(as.Date(all_trips$date),"%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date),"%A")

```

add a ride length calculation to all_trips(in seconds)

```{r calculating ride_length}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
str(all_trips)
```
convert ride_length from factor to numeric so we can run calculations on data

```{r converting ride_length to numeric}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

Remove "bad" data, The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative. We will create a new version of the dataframe (v2) since data is being removed

```{r removing bad data}
all_trips_V2 <- all_trips[!(all_trips$start_station_name=="HQ QR"|all_trips$ride_length<0),]

```

## Conducting Descriptive Analysis

Descriptive analysis on ride_length (all figures in seconds).

```{r conducting descriptive analysis1}
mean(all_trips_V2$ride_length)
median(all_trips_V2$ride_length)
min(all_trips_V2$ride_length)
max(all_trips_V2$ride_length)
summary(all_trips_V2$ride_length)
```
comparing members and casual Users, based on their ride_length

```{r comparing members and casual users}
aggregate(all_trips_V2$ride_length ~ all_trips_V2$member_casual, FUN = mean)
aggregate(all_trips_V2$ride_length ~ all_trips_V2$member_casual, FUN = median)
aggregate(all_trips_V2$ride_length ~ all_trips_V2$member_casual, FUN = max)
aggregate(all_trips_V2$ride_length ~ all_trips_V2$member_casual, FUN = min)
```
noticed that the days of the week are are out of order let's fix that

```{r fixing order of day of the week column}
all_trips_V2$day_of_week <- ordered(all_trips_V2$day_of_week, levels = c('Sunday','Monday','Tuesday','Wednesday','Thursday','Friday','Saturday'))

```


let's run the average ride time by each day for members vs casual users

```{r average ride time by each day}
aggregate(all_trips_V2$ride_length ~ all_trips_V2$member_casual + all_trips_V2$day_of_week, FUN = mean)

```

analyzing and visualizing the number of rides by rider_type

```{r analyzing and visualizing the number of rides by rider type}
all_trips_V2 %>% 
  mutate(weekday = wday(started_at,label = TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x=weekday,y=number_of_rides,fill=member_casual))+geom_col(position='dodge')
```

creating a visualization for average duration

```{r visualization for average duration}
all_trips_V2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(number_of_rides = n(),average_duration=mean(ride_length)) %>% 
  arrange(member_casual,weekday) %>% 
  ggplot(aes(x=weekday,y=average_duration,fill=member_casual))+geom_col(position='dodge')
```

visualizing the data by number of rides by member_casual and month

```{r visualizing number of rides by month}
all_trips_V2 %>% 
  group_by(member_casual,month) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual,month) %>% 
  ggplot(aes(x=month,y=number_of_rides,fill=member_casual))+geom_col(position='dodge')

```
visualizing the data by avearge duration of rides by member_casual and month

```{r visualizing average duration by month}
all_trips_V2 %>% 
  group_by(member_casual,month) %>% 
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>% 
  arrange(member_casual,month) %>% 
  ggplot(aes(x=month,y=average_duration,fill=member_casual))+geom_col(position='dodge')

```




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
