---
title: "CyclisticBikeShare"
author: "Tejaswi"
date: "2022-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This research is for Google Data Analytics Certificate Capstone Project.It was inspired by Kevin Hartman's case study "'Sophisticated, Clear, and Polished': Divvy and Data Visualization" (available here:https://artscience.blog/home/divvy-dataviz-case-study).

## Scenario

I am a junior data analyst working in the marketing analyst team at Cyclistic, a bike share company in chicago. The director of marketing believes the company's future success depends on maximizing the number of annual memberships. Therefore my team wants to understand how casual riders and annual members use cyclistic bikes differently and design a new marketing strategy to convert casual riders into annual members

## Key Stakeholders

* Lily Moreno, Director of Marketing
* Marketing Analytics Team
* Cyclistic Executive Team

## Business Task

To understand how casual riders and annual members use cyclistic bikes differently and design marketing strategy to maximize the number of annual Memberships by converting casual riders into annual members

## Preparing Data

The data has been made available by Motivate International Inc. The data is original and [Licensed](https://ride.divvybikes.com/data-license-agreement)

Citation - The data is available on https://www.divvybikes.com/system-data and is owned by the City of Chicago

#### Installing required Packages

Use install.packages("tidyverse") to install tidyverse
Tidyverse - A system of packages in R with a common design philosophy for data manipulation, exploration and visualization

Lubridate - It is an R package that makes it easier to work with dates and times

ggplot2 - Creates a variety of data visualizations by applying different visual properties to the data variables in R.

```{r cars}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

#### Collecting Data

readr - It is used for importing data

read_csv() -> This function is used to import csv files and reads the data as data frames from it. it is available under readr package.
```{r collecting data}
library(readr)
# Uploading divvy datasets here
Q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
Q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
Q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
Q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

#### Wrangle Data and Combine to a Single File

compare column names of each of the files
While the column names need not be in same order, but they do need to perfectly match before we can use a command to join them into one file.

colnames() - it is a function in R used to obtain the names of columns in a matrix-like object.
```{r wrangle data}
colnames(Q2_2019)
colnames(Q3_2019)
colnames(Q4_2019)
colnames(Q1_2020)
```

Rename columns to make them consistent with Q1_2020(as this will be the supposed going forward table design for Divvy)
to rename the columns we use rename method.

rename() - it is a method available in the dplyr package, which is used to change the particular column name present in the data frame. The operator - %>% (pipe) is used to load the renamed column names to the data frame. At a time we can change single or multiple column names

str() - it is a function in R language used for compactly displaying the internal structure of a R object.
```{r renaming columns}
Q4_2019 <- rename(Q4_2019,
                  ride_id = trip_id,
                  rideable_type = bikeid,
                  started_at = start_time,
                  ended_at = end_time,
                  start_station_name = from_station_name,
                  start_station_id = from_station_id,
                  end_station_name = to_station_name,
                  end_station_id = to_station_id,
                  member_casual = usertype)
(Q3_2019 <- rename(Q3_2019,
                   ride_id = trip_id,
                   rideable_type = bikeid,
                   started_at = start_time,
                   ended_at = end_time,
                   start_station_name = from_station_name,
                   start_station_id = from_station_id,
                   end_station_name = to_station_name,
                   end_station_id = to_station_id,
                   member_casual = usertype))
Q2_2019 <- rename(Q2_2019,
                  ride_id = "01 - Rental Details Rental ID",
                  rideable_type = "01 - Rental Details Bike ID",
                  started_at = "01 - Rental Details Local Start Time",
                  ended_at = "01 - Rental Details Local End Time",
                  start_station_name = "03 - Rental Start Station Name",
                  start_station_id = "03 - Rental Start Station ID",
                  end_station_name = "02 - Rental End Station Name",
                  end_station_id = "02 - Rental End Station ID",
                  member_casual = "User Type")
str(Q2_2019)
str(Q3_2019)
str(Q4_2019)
str(Q1_2020)
```
Convert ride_id and rideable_type to character so that they can stack correctly

mutate() - it adds new variables while retaining old variables to a data frame.
```{r converting data type}
Q4_2019 <- mutate(Q4_2019,
                  ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
Q3_2019 <- mutate(Q3_2019,
                  ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
Q2_2019 <- mutate(Q2_2019,
                  ride_id = as.character(ride_id),rideable_type = as.character(rideable_type))
```

stack individual quarters data frames to one data frame

bind_rows() - it is a function used to combine rows of two data frames.

select() - it is used to pick specific variables or features of a data frame or a tibble
```{r stacking to one data frame}
all_trips <- bind_rows(Q2_2019,Q3_2019,Q4_2019,Q1_2020)
all_trips <- all_trips %>% 
  select(-c(start_lat,start_lng,end_lat,end_lng,birthyear,gender,"01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year","Member Gender","tripduration"))
```

## clean up and add data for analysis

inspect the new table that has been created

nrow() - used to return the number of rows of the specified data frame or matrix.

dim() - used to set or get the dimensions of the specified matrix, array or data frame.

head() - used to display the first n rows present in the input data frame, by default it will show first 6 rows of data frame.

summary() - to produce or display the summary of all records present in the found set.
```{r inspecting data}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)

```
There are few problems that need to be fixed: they are

* In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.

* The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.

* We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.

* There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.


 In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual". Before 2020, Divvy used different labels for these two types of riders ... we will want to make our dataframe consistent with their current nomenclature
 
```{r renaming values of member_casual}
table(all_trips$member_casual)
all_trips <- all_trips %>% 
  mutate(member_casual = recode(member_casual,"Subscriber" = "member", "Customer" = "casual"))
table(all_trips$member_casual)

```

Add columns that list the date, month, day, and year of each ride. This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level

format() - it is used to format strings, dates and numbers in a specified style.

```{r adding columns that list date,month,year and date}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date),"%m")
all_trips$day <- format(as.Date(all_trips$date),"%d")
all_trips$year <- format(as.Date(all_trips$date),"%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date),"%A")

```

add a ride length calculation to all_trips(in seconds)

difftime() - it computes the difference between two objects of date or time.
```{r calculating ride_length}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
str(all_trips)
```
convert ride_length from factor to numeric so we can run calculations on data

```{r converting ride_length to numeric}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

Remove "bad" data, The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative. We will create a new version of the dataframe (v2) since data is being removed

```{r removing bad data}
all_trips_V2 <- all_trips[!(all_trips$start_station_name=="HQ QR"|all_trips$ride_length<0),]

```

## Conducting Descriptive Analysis

Descriptive analysis on ride_length (all figures in seconds).

mean() - used to calculate the arithmetic mean of the elements of the numeric vector passed to it as argument

median() - it accepts a vector as an input . if there are an odd number of values in the vector , the function returns the middle valueif there are even number of values in the vector , the function returns the average of two medians

min() - used to find the minimum element present in an object. this object can be a vector, a list, a matrix, a data frame etc.

max() -  used to find the maximum element present in an object. this object can be a vector, a list, a matrix, a data frame etc.

```{r conducting descriptive analysis1}
mean(all_trips_V2$ride_length)
median(all_trips_V2$ride_length)
min(all_trips_V2$ride_length)
max(all_trips_V2$ride_length)
summary(all_trips_V2$ride_length)
```
comparing members and casual Users, based on their ride_length

aggregate() - it is used to get the summary statistics of the data by group
```{r comparing members and casual users}
aggregate(all_trips_V2$ride_length ~ all_trips_V2$member_casual, FUN = mean)
aggregate(all_trips_V2$ride_length ~ all_trips_V2$member_casual, FUN = median)
aggregate(all_trips_V2$ride_length ~ all_trips_V2$member_casual, FUN = max)
aggregate(all_trips_V2$ride_length ~ all_trips_V2$member_casual, FUN = min)
```
noticed that the days of the week are are out of order let's fix that

```{r fixing order of day of the week column}
all_trips_V2$day_of_week <- ordered(all_trips_V2$day_of_week, levels = c('Sunday','Monday','Tuesday','Wednesday','Thursday','Friday','Saturday'))

```


let's run the average ride time by each day for members vs casual users

```{r average ride time by each day}
aggregate(all_trips_V2$ride_length ~ all_trips_V2$member_casual + all_trips_V2$day_of_week, FUN = mean)

```

analyzing and visualizing the number of rides by rider_type

group_by() - it belongs to the dplyr package in R, which groups the data frames

summarise() - it is used in the R program to to summarize the data frame into just one value or vector. This summarization is done through grouping observations by using categorical values at first, using the group_by() function. the dplyr package is used to get the summary of dataset.

arrange() - it is a function used in R to reorder the rows of a data frame/table by using column names.

ggplot() - it is a plotting package that provides helpful commands to create complex plots from data in a date frame.

```{r analyzing and visualizing the number of rides by rider type}
all_trips_V2 %>% 
  mutate(weekday = wday(started_at,label = TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x=weekday,y=number_of_rides,fill=member_casual))+geom_col(position='dodge')
```

creating a visualization for average duration

```{r visualization for average duration}
all_trips_V2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(number_of_rides = n(),average_duration=mean(ride_length)) %>% 
  arrange(member_casual,weekday) %>% 
  ggplot(aes(x=weekday,y=average_duration,fill=member_casual))+geom_col(position='dodge')
```

visualizing the data by number of rides by member_casual and month

```{r visualizing number of rides by month}
all_trips_V2 %>% 
  group_by(member_casual,month) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual,month) %>% 
  ggplot(aes(x=month,y=number_of_rides,fill=member_casual))+geom_col(position='dodge')

```
visualizing the data by avearge duration of rides by member_casual and month

```{r visualizing average duration by month}
all_trips_V2 %>% 
  group_by(member_casual,month) %>% 
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>% 
  arrange(member_casual,month) %>% 
  ggplot(aes(x=month,y=average_duration,fill=member_casual))+geom_col(position='dodge')

```

## Summary of above analysis

*  In a week, maximum number of rides are taken by annual members when compared with number of rides taken by casual members

*  In a week less rides are taken by annual members on weekend(i.e saturday and sunday) when compared with other days

*  In a week more rides are taken by casual members on weekend(i.e saturday and sunday) when compared with other days

*  In a week average ride length of casual members is greater than average ride length of annual members

*  In a year number of rides are higher in mid-year(april,may,june,july,august,september,october) by both annual members and casual members

*  In a year the average duration of rides is greater in january and decreases gradually in further months. especially for casual members.


## Recommendations based on my analysis

*  As casual customers average ride_length is high, could be that most of them are travelling between far places. So, using this some kind of offers or schemes could be introduced to attract those casual customers to avail annual membership and provide them some benefits as they travel for long distance

ex: if they have this annual membership, they could save some expense on ride between two far places upto xkm/m

(or)

if the casual customer is using the drives on daily basis, they could be charged with a reasonable monthly charge 
rather than daily charge as it would save expense of customer on daily rides.

*  On weekends, and holiday seasons casual members take more number of rides, so we could introduce some benefits on rides taken on weekends for only annual members, this would help in attracting casual customers to take the annual membership.

* Making sure to spread the awareness of benefits and facilities of annual membership.

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
